\documentclass[11pt]{article}
\usepackage[review]{acl}
\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{makecell}
\usepackage{enumitem}
\usepackage{xcolor}

\setlength\titlebox{6.5cm}

\newcommand{\method}{\textsc{INSIGHT}}
\newcommand{\methodfull}{Interpretable Safety-Guided Accident Anticipation via Concept Bottleneck and Temporal Actor-Critic}
\newcommand{\cbm}{\textsc{CBM}}
\newcommand{\caac}{\textsc{CAAC}}
\newcommand{\cgta}{\textsc{CGTA}}
\newcommand{\crs}{\textsc{CRS}}
\newcommand{\tsd}{\textsc{TSD}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\etal}{et~al.}

\title{\method: A Language-Grounded Risk Concept Interface for Interpretable Accident Anticipation}

\author{Anonymous Submission}

\begin{document}
\maketitle

\begin{abstract}
Traffic accident anticipation models are usually evaluated as opaque binary
predictors: given a video prefix, estimate whether an accident will occur.
For safety-critical use, the missing object is not only a better score, but an
auditable semantic interface. Such an interface should expose what the model
treats as risky, make alternative vocabularies comparable, and let researchers
ask how semantic evidence affects anticipation behavior. Current methods rarely
offer this object; most optimize prediction quality while leaving their
internal evidence latent or relying on post-hoc explanation.

We present \method{}, a framework for interpretable accident anticipation built
around a governed language-grounded risk concept interface. The core
contribution is a reproducible multimodal concept construction pipeline that
mines candidate risk concepts from driving frames, performs risk-first
refinement, applies family balancing and canonical merge consolidation, and
exports a trainable ontology with explicit review metadata, family structure,
and merge provenance. This ontology becomes the semantic bottleneck of the
anticipation model rather than an appendix-only descriptive artifact.

On top of this interface, \method{} trains a concept-bottleneck anticipation
model with concept-linked temporal modeling and auditable per-concept scoring.
This design lets us study not only prediction quality, but also how ontology
choice changes the predictive--timing operating point under matched launchers.
Across DAD and A3D, matched ontology comparisons show that ontology source
materially affects the AP--mTTA frontier, making semantic interface design a
first-class modeling decision rather than hidden preprocessing. We pair these
model-side results with language-side audits over a 500-frame DAD slice,
canonical naming and paraphrase robustness checks, and light human review
across all 80 released concepts, so that the ontology can be evaluated as a
governed semantic artifact rather than as a prompt byproduct. The resulting model
reaches a strong interpretable operating point, including \textbf{93.40\% AP}
and \textbf{4.90\,s mTTA} on A3D, and \textbf{68.19\% AP} on the canonical DAD
curriculum line, while using only $\sim$30M parameters. We further present
semantic audit and intervention analyses showing that the concept layer is
structurally meaningful and partially intervenable; actor-policy timing is
reported as scoped supporting evidence rather than as the main claim.
\end{abstract}

\input{sec_intro_emnlp}
\input{sec_related_emnlp}
\input{sec_method_emnlp}
\input{sec_experiments_emnlp}
\input{sec_conclusion_emnlp}

\clearpage
\appendix
\input{../neurips2026/sec_appendix}

\clearpage
\bibliography{insight}

\end{document}
