\documentclass{article}
\usepackage{neurips_2026}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{makecell}
\usepackage{enumitem}
\usepackage{ulem}

% ── Caption / float spacing ────────────────────────────────────────────────
\captionsetup{font=small,labelfont=bf,singlelinecheck=false}
\setlength{\textfloatsep}{10pt plus 2pt minus 2pt}
\setlength{\floatsep}{8pt plus 2pt minus 2pt}
\setlength{\intextsep}{8pt plus 2pt minus 2pt}

% ── Colours (inspired by mm25 CAMERA paper) ────────────────────────────────
\definecolor{accentblue}{RGB}{31,119,180}
\definecolor{accentorange}{RGB}{255,127,14}
\definecolor{accentgreen}{RGB}{44,160,44}
\definecolor{accentred}{RGB}{214,39,40}

% ── Commands ────────────────────────────────────────────────────────────────
\newcommand{\method}{\textsc{INSIGHT}}
\newcommand{\methodfull}{Interpretable Safety-Guided Accident Anticipation
  via Concept Bottleneck and Temporal Actor-Critic}
\newcommand{\cbm}{\textsc{CBM}}
\newcommand{\caac}{\textsc{CAAC}}
\newcommand{\cgta}{\textsc{CGTA}}
\newcommand{\crs}{\textsc{CRS}}
\newcommand{\tsd}{\textsc{TSD}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\etal}{et~al.}

\title{\method: Interpretable Accident Anticipation with Concept-Guided Risk Reasoning and Timing}

\author{Anonymous Authors \\ Anonymous Institution}

\begin{document}
\maketitle

\begin{abstract}
Traffic accident anticipation is usually framed as a binary prediction task:
given a video prefix, estimate whether an accident will occur.
We argue that this framing is incomplete.
A deployment-ready anticipation system must address two coupled questions:
what semantic evidence makes the scene dangerous, and how can that evidence be
used to produce earlier warnings rather than merely later confidence.
Current methods address these needs only partially: most optimize
classification accuracy rather than warning timeliness, and almost all remain
black boxes or rely on post-hoc explanation.
We therefore reformulate accident anticipation as a
\emph{concept-conditioned sequential decision problem}.

We present \method{} (\textbf{In}terpretable \textbf{S}afety-\textbf{G}uided
accident ant\textbf{i}cipation via Concept Bott\textbf{le}neck and
Concep\textbf{t}-aware Actor-Critic), a framework with two tightly coupled
parts. A Concept Bottleneck Module (\cbm) maps per-frame object features to
human-interpretable risk concepts, providing the paper's strongest and most
directly supported interpretability layer. A concept-conditioned timing module
then uses these concepts inside the decision state, supplying an explicit route
from semantic evidence to warning control, although the current paper's
policy-level evidence is more limited than its WHY-layer evidence. Three
additional components---Concept-Guided Temporal Attention (\cgta), Concept Risk
Score (\crs), and Temporally Shifted Distillation (\tsd)---improve temporal
focus, auditability, and anticipatory representation learning.

Beyond the architecture, we introduce a reproducible multimodal risk concept
construction pipeline that mines candidate concepts from driving frames,
performs risk-first refinement, applies family balancing and canonical merge
consolidation, and yields an auditable ontology with explicit family metadata
and merge provenance. This turns the concept bottleneck from a fixed handcrafted
vocabulary into a reusable semantic interface and allows matched ontology
comparisons under a shared launcher protocol.

On A3D, \method{} achieves \textbf{93.40\% AP}, \textbf{mTTA\,=\,4.90\,s},
\textbf{TTA@R80\,=\,4.89\,s}, and \textbf{P@R80\,=\,0.9067}.
On DAD, \method{} achieves \textbf{68.19\% AP} under the headline curriculum
protocol and remains competitive within the interpretable comparison tier.
Matched score-source comparisons on representative checkpoints further show
that actor-based timing remains substantially weaker than classifier-based
anticipation on DAD, but is much closer on A3D. We also show that ontology
choice changes the AP--mTTA operating point under matched launchers, and that
archived intervention analyses provide quantitative but still partial support
for a concept-conditioned timing story. All results are obtained with only
$\sim$30M parameters.
\end{abstract}

\input{sec_intro}
\input{sec_related}
\input{sec_method}
\input{sec_experiments}
\input{sec_conclusion}

\clearpage
\appendix
\input{sec_appendix}

\clearpage
\bibliographystyle{plainnat}
\bibliography{insight}

\clearpage
\input{checklist}

\end{document}
